---
title: "[R] Training: Beginners to Advanced - Statistics"
output:
  html_document:
    df_print: paged
---

# Topic 4 - Statistics
This topic covers a basic introduction to statistics in R.  There are some very comprehensive statistical packages available depending on your needs.  Some links to these are provided in the notes.

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=48),tidy=TRUE)
```

## Example 1: Calculation of descriptive statistics
We will use the trees datyaset that comes with R. Descriptive statistics just provide a quick overview of the data.
```{r}
data(trees)
head(trees)
summary(trees)
mean(trees$Girth)
mean(trees$Height)
sd(trees$Girth)
median(trees$Girth)
library(lattice)
histogram(trees$Girth)
plot(x=trees$Girth,y=trees$Volume)  # looks like some kind of relationship here...data
cor(trees$Girth,trees$Volume)
```


## Example 3: Linear Regression
From the plot above, it looks like there could be a linear relationship between Girth and Volume.  Lets find out.
```{r}
# lets draw it up with ggplot
library(ggplot2)
ggplot(data=trees,aes(x=Girth,y=Volume)) + geom_point() + geom_smooth(method="lm") + ggtitle("Volume and Girth relationship") 

myTreeRegression <- lm(data = trees, formula = Volume ~ Girth)
myTreeRegression
summary(myTreeRegression)
```

